---
title: "mh_basidia"
output: html_document
date: "2026-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyze basidia data

This initial analysis is to see if there are really some glaring data deficiencies that we should shore up with a bit more microscopy. So, I am going to make a few figures show the coverage of microscopy across the genus

```{r}
setwd("~/Desktop/wiscam_microscopy/data/mh_basidia")
```

Making a section newick tree to use in visualizations
```{r}
library(ape)
library(nlme)
library(phytools)

sectree <- read.tree('sec_tre.nwk')
plot(sectree)

#Rotate nodes
desired_order <- c(
  "Lepidella",
  "Roanokensis",
  "Amidella",
  "Strobiloformes",
  "Validae",
  "Phalloideae",
  "Arenariae",
  "Amarrendiae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)
desired_order <- rev(desired_order)

sectree_ordered <- rotateConstr(sectree, desired_order)
plot(sectree_ordered)
```

Total basidia counted graph:
```{r}
library(tidyverse)
library(ggplot2)
library(ggtree)
library(patchwork)
library(tidyr)

basidia <- read_csv("basidia_filled_mh.csv", show_col_types = FALSE)

#Identify sterigma count columns
ster_cols <- c("4s/basidia", "3s/basidia", "2s/basidia", "s/basidia")

#Calculate total basidia observed for each specimen
basidia_totals <- basidia %>%
  mutate(
    total_basidia = rowSums(across(all_of(ster_cols)), na.rm = TRUE)
  ) %>%
  filter(total_basidia > 0)

#Sum total basidia observed per section
section_totals <- basidia_totals %>%
  group_by(sectionid) %>%
  summarise(
    total_basidia = sum(total_basidia),
    n_specimens = n(),
    .groups = "drop"
  )

section_totals <- section_totals %>%
  mutate(sectionid = as.character(sectionid))

#Add missing sections
tree_sections <- sectree_ordered$tip.label

section_totals_full <- section_totals %>%
  complete(
    sectionid = tree_sections,
    fill = list(total_basidia = 0, n_specimens = 0)
  )

#Order phylogenetically
phylo_order <- c(
  "Lepidella",
  "Roanokensis",
  "Amidella",
  "Strobiloformes",
  "Validae",
  "Phalloideae",
  "Arenariae",
  "Amarrendiae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

#Plot section totals
basidia_count <- ggplot(section_totals_full,
                aes(x = sectionid, y = total_basidia, fill = sectionid)) +
  geom_col(color = "black") +
  geom_text(
    aes(label = ifelse(total_basidia > 0, total_basidia, "")),
    vjust = -0.3,
    size = 3
  ) +
  scale_x_discrete(limits = phylo_order) +
  labs(
    x = "Amanita Section",
    y = "Total basidia observed"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Now visualizing sterigma counts per basidia for each specimen organized by section phylogenetically
```{r}
library(tidyverse)

#Graph order
phylo_order <- c(
  "Lepidella",
  "Roanokensis",
  "Amidella",
  "Strobiloformes",
  "Validae",
  "Phalloideae",
  "Arenariae",
  "Amarrendiae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

#Input data
basidia <- read_csv("basidia_filled_mh.csv", show_col_types = FALSE)
ster_cols <- c("4s/basidia", "3s/basidia", "2s/basidia", "s/basidia")
basidia_clean <- basidia %>%
  filter(if_any(all_of(ster_cols), ~ !is.na(.)))

#Converts sterigma count to per-basidium observations
ster_long <- basidia_clean %>%
  select(specimen_id, species, sectionid, all_of(ster_cols)) %>%
  pivot_longer(
    cols = all_of(ster_cols),
    names_to = "sterigma_class",
    values_to = "count"
  ) %>%
  filter(!is.na(count), count > 0) %>%
  mutate(
    sterigmata = case_when(
      sterigma_class == "4s/basidia" ~ 4,
      sterigma_class == "3s/basidia" ~ 3,
      sterigma_class == "2s/basidia" ~ 2,
      sterigma_class == "s/basidia"  ~ 1
    )
  )

ster_expanded <- ster_long %>%
  uncount(weights = count)
#Organize phylogenetically

ster_expanded$sectionid <- factor(
  ster_expanded$sectionid,
  levels = phylo_order
)

#Plot showing the variablility of sterigmata counts throught the sections
dot_chart <- ggplot(
  ster_expanded,
  aes(x = specimen_id, y = sterigmata)
) +
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.4,
    size = 0.6
  ) +
  facet_wrap(
    ~ sectionid,
    scales = "free_x",
    nrow = 1
  ) +
  scale_y_continuous(breaks = 1:4) +
  labs(
    x = "Specimen",
    y = "Sterigmata per basidium",
    title = "Sterigma number variation per specimen",
    subtitle = "Specimens grouped and ordered by Amanita section"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_blank()
  )

```

Making a bar chart that aggregates the data to each species and stacks sterigma count for each species
```{r}
library(tidyverse)

basidia <- read_csv("basidia_filled_mh.csv", show_col_types = FALSE)

ster_cols <- c("4s/basidia", "3s/basidia", "2s/basidia", "s/basidia")

basidia_clean <- basidia %>%
  filter(if_any(all_of(ster_cols), ~ !is.na(.)))

#Aggregate based on species
species_sterigma <- basidia_clean %>%
  group_by(sectionid, species) %>%
  summarise(
    `4` = sum(`4s/basidia`, na.rm = TRUE),
    `3` = sum(`3s/basidia`, na.rm = TRUE),
    `2` = sum(`2s/basidia`, na.rm = TRUE),
    `1` = sum(`s/basidia`,  na.rm = TRUE),
    .groups = "drop"
  )

#Convert to long format

species_long <- species_sterigma %>%
  pivot_longer(
    cols = c(`4`, `3`, `2`, `1`),
    names_to = "sterigmata",
    values_to = "count"
  ) %>%
  filter(count > 0)

#Organize phylogenetically
phylo_order <- c(
  "Lepidella",
  "Roanokensis",
  "Amidella",
  "Strobiloformes",
  "Validae",
  "Phalloideae",
  "Arenariae",
  "Amarrendiae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

species_long$sectionid <- factor(
  species_long$sectionid,
  levels = phylo_order
)

# Order species within sections
species_long <- species_long %>%
  mutate(
    species = fct_reorder(
      species,
      as.numeric(sectionid)
    )
  )

#Plot
stercount_plot <- ggplot(
  species_long,
  aes(x = species, y = count, fill = sterigmata)
) +
  geom_col(color = "black", linewidth = 0.2) +
  facet_grid(
    ~ sectionid,
    scales = "free_x",
    space = "free_x"
  ) +
  scale_fill_brewer(
    palette = "Set2",
    name = "Sterigmata\nper basidium"
  ) +
  labs(
    x = "Species",
    y = "Number of basidia observed",
    title = "Sterigma number composition per species",
    subtitle = "Species grouped by Amanita section (phylogenetic order)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    strip.background = element_blank()
  )
##To save the plot: ggsave(filename = stericount_species.png, plot = stericount_plot, width = 10, height = 6, units = "in", dpi = 300)
```


Now let's check out species aggregations
```{r}

# If your CSV uses slightly different column names, standardize them here:
# - species column -> 'species'
# - basidia count  -> 'basidia_count'
df_spec <- df0 %>%
  rename(
    species        = coalesce(names(.)[str_detect(names(.), regex("^species$", ignore_case = TRUE))][1], "species"),
    basidia_count  = coalesce(names(.)[str_detect(names(.), regex("^basidia",  ignore_case = TRUE))][1], "basidia_count")
  ) %>%
  mutate(basidia_count = readr::parse_number(as.character(basidia_count))) %>%
  filter(!is.na(basidia_count))
# Trying to pull section data into separate datasets

```


```{r}
library(readr)
library(dplyr)
library(ggplot2)

# Read your CSV
basidia <- read_csv("basidia_filled_rt.csv")

# Filter to specimens with BOTH avg_length and avg_width
basidia_clean <- basidia %>%
  filter(!is.na(avg_length), !is.na(avg_width))

# Define your color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Scatter plot
ggplot(basidia_clean, aes(x = avg_width, y = avg_length, color = sectionid)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Average Basidia Dimensions by Specimen",
    x = "Average Basidia Width (µm)",
    y = "Average Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14)

```

```{r}
#making a box and whisker plot of all the Vaginatae basidia lengths

library(dplyr)
library(ggplot2)
library(stringr)

# Start with your dataset
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to Vaginatae with complete basidia length data
vag <- basidia %>%
  filter(sectionid == "Vaginatae",
         !is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Create formatted species labels:
# Convert "Amanita fusca" → "A. fusca"
# Keep "Amanita sp aurorae" → "A. sp aurorae"
vag <- vag %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A."),
    species_short = factor(species_short, levels = unique(species_short))
  )

# Build the plot
ggplot(vag, aes(x = species_short)) +
  # Whisker line from min to max
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = species_short), linewidth = 1) +
  # Point for average
  geom_point(aes(y = avg_length), size = 3, color = "black") +
  labs(
    title = "Basidia Length Range per Specimen (Vaginatae Only)",
    x = "Species",
    y = "Basidia Length (µm)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length data
all_clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Create species labels: "Amanita fusca" → "A. fusca"
all_clean <- all_clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort by section, then species, then specimen_id
all_clean <- all_clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(all_clean, aes(x = specimen_id, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = specimen_id), linewidth = 1) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  scale_x_discrete(labels = all_clean$species_short) +
  labs(
    title = "Basidia Length Range per Specimen (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )
```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length data
all_clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Format species names: "Amanita fusca" → "A. fusca"
all_clean <- all_clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Reorder species by section first, then species name
all_clean <- all_clean %>%
  arrange(sectionid, species_short) %>%
  mutate(species_short = factor(species_short, levels = unique(species_short)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(all_clean, aes(x = species_short, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = species_short), linewidth = 1) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Basidia Length Range per Specimen (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

```

```{r}
#the average of the average within each section, the average minimum, the average maximum
library(dplyr)
library(ggplot2)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to rows with complete basidia length data
clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Summarize by section
section_summary <- clean %>%
  group_by(sectionid) %>%
  summarise(
    mean_min = mean(min_length, na.rm = TRUE),
    mean_avg = mean(avg_length, na.rm = TRUE),
    mean_max = mean(max_length, na.rm = TRUE)
  ) %>%
  arrange(sectionid)

# Color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(section_summary, aes(x = sectionid, color = sectionid)) +
  # Whisker from mean min to mean max
  geom_segment(aes(y = mean_min, yend = mean_max,
                   xend = sectionid), linewidth = 1.2) +
  # Point for mean average
  geom_point(aes(y = mean_avg), size = 4) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Average Basidia Length Range per Amanita Section",
    x = "Section",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)
install.packages("patchwork")
library(patchwork)   # for combining plots

# Read data
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length & width
clean <- basidia %>%
  filter(!is.na(avg_length),
         !is.na(avg_width),
         !is.na(min_length),
         !is.na(max_length))

# Format species names: "Amanita fusca" → "A. fusca"
clean <- clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort specimens by section, then species, then specimen_id
clean <- clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# -----------------------------
# Panel A: Scatterplot
# -----------------------------
p1 <- ggplot(clean, aes(x = avg_width, y = avg_length, color = sectionid)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "A. Scatterplot: Basidia Avg Width vs Avg Length",
    x = "Average Basidia Width (µm)",
    y = "Average Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14)

# -----------------------------
# Panel B: Specimen Whiskers
# -----------------------------
p2 <- ggplot(clean, aes(x = specimen_id, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = specimen_id), linewidth = 1.2) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  scale_x_discrete(labels = clean$species_short) +
  labs(
    title = "B. Specimen Basidia Length Ranges (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

# -----------------------------
# Combine Panels
# -----------------------------
p1 / p2   # stacked vertically

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

# Read data
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with basidia_count
clean <- basidia %>%
  filter(!is.na(basidia_count))

# Create species labels: "Amanita fusca" → "A. fusca"
clean <- clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort by section, then species, then specimen_id
clean <- clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Bar graph
ggplot(clean, aes(x = specimen_id, y = basidia_count, fill = sectionid)) +
  geom_col(color = "black") +
  scale_fill_manual(values = section_colors) +
  scale_x_discrete(labels = clean$species_short) +
  labs(
    title = "Number of Basidia Measured per Specimen",
    x = "Species",
    y = "Basidia Count",
    fill = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )




