---
title: "mh_basidia"
output: html_document
date: "2026-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyze basidia data

This initial analysis is to see if there are really some glaring data deficiencies that we should shore up with a bit more microscopy. So, I am going to make a few figures show the coverage of microscopy across the genus

```{r}
setwd("~/Desktop/wiscam_microscopy/data/mh_basidia")
```

Loading basidia data and making a plot of basidia count per section

```{r}
# install.packages("tidyverse")   # run once if needed
library(tidyverse)

# === 1) Load your CSV ========================================================
# Replace the path/filename as needed
df0 <- readr::read_csv("basidia_filled_mh.csv", show_col_types = FALSE)

# If your CSV uses different column names, standardize them here
#   - section column: sectionid (or Section, etc.)
#   - basidia count column: basidia_count (or BasidiaCount, etc.)
df <- df0 %>%
  rename(
    sectionid     = coalesce(names(.)[str_detect(names(.), regex("^section", ignore_case = TRUE))][1], "sectionid"),
    basidia_count = coalesce(names(.)[str_detect(names(.), regex("^basidia", ignore_case = TRUE))][1], "basidia_count")
  )

# Ensure numeric basidia_count
df <- df %>%
  mutate(basidia_count = readr::parse_number(as.character(basidia_count))) %>%
  filter(!is.na(basidia_count))

# === 2) Define 7 target sections =======================================
sections_7 <- c(
  "Phalloideae",
  "Vaginatae",
  "Amanita",
  "Validae",
  "Roanokensis",
  "Caesareae",
  "Amidella"
)

# If your raw data has variants/synonyms you want to fold into these 7,
# add them in case_when() below. If not, keep the simple default mapping.
df <- df %>%
  mutate(
    section_raw = as.character(sectionid),

    # ----- EDIT THIS MAPPING as needed -----
    # Put any synonyms/variants on the left (using regex or exact strings),
    # and map them to one of the 7 names from sections_7 on the right.
    section_7 = case_when(
      str_detect(section_raw, regex("^phallo", ignore_case = TRUE)) ~ "Phalloideae",
      str_detect(section_raw, regex("^vagin",  ignore_case = TRUE)) ~ "Vaginatae",
      str_detect(section_raw, regex("^amanita$", ignore_case = TRUE)) ~ "Amanita",
      str_detect(section_raw, regex("^valida", ignore_case = TRUE)) ~ "Validae",
      str_detect(section_raw, regex("^roanok", ignore_case = TRUE)) ~ "Roanokensis",
      str_detect(section_raw, regex("^caesar", ignore_case = TRUE)) ~ "Caesareae",
      str_detect(section_raw, regex("^amidell", ignore_case = TRUE)) ~ "Amidella",

      # Anything else -> "Other/Unassigned" (optional; you can drop instead)
      TRUE ~ "Other/Unassigned"
    )
  ) %>%
  # Keep only the 7 sections (drop the rest) — OR comment this line if you want an "Other" bar
  filter(section_7 %in% sections_7) %>%
  # Lock the display order to your 7 names (use this order in charts)
  mutate(section_7 = factor(section_7, levels = sections_7))

# === 3) Aggregate totals per section (bar chart) =============================
by_section <- df %>%
  group_by(section_7) %>%
  summarise(
    total_basidia       = sum(basidia_count, na.rm = TRUE),
    specimens           = sum(!is.na(basidia_count)),
    mean_per_specimen   = total_basidia / pmax(specimens, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(total_basidia))

# ---- BAR CHART: total basidia per section ----
p_totals <- ggplot(by_section, aes(x = section_7, y = total_basidia)) +
  geom_col(fill = "#3B82F6") +
  geom_text(aes(label = total_basidia), vjust = -0.2, size = 3.5) +
  labs(
    title = "Total basidia per section",
    x = "Section",
    y = "Total basidia counted"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

p_totals

# Save if desired
# ggsave("basidia_total_per_section.png", p_totals, width = 8, height = 5, dpi = 300)

# === 4) Optional: histogram of per-specimen counts, faceted by section =======
# This shows within-section distributions (not aggregated)
p_hist <- ggplot(df, aes(x = basidia_count)) +
  geom_histogram(binwidth = 1, boundary = 0, color = "white", fill = "#10B981") +
  facet_wrap(~ section_7, ncol = 3, scales = "free_y") +
  labs(
    title = "Distribution of basidia counts by section",
    x = "Basidia per specimen",
    y = "Number of specimens"
  ) +
  theme_minimal(base_size = 12)

# Print if you want to inspect
# p_hist
# ggsave("basidia_hist_by_section.png", p_hist, width = 10, height = 7, dpi = 300)

```

Now let's check out species aggregations
```{r}

# If your CSV uses slightly different column names, standardize them here:
# - species column -> 'species'
# - basidia count  -> 'basidia_count'
df_spec <- df0 %>%
  rename(
    species        = coalesce(names(.)[str_detect(names(.), regex("^species$", ignore_case = TRUE))][1], "species"),
    basidia_count  = coalesce(names(.)[str_detect(names(.), regex("^basidia",  ignore_case = TRUE))][1], "basidia_count")
  ) %>%
  mutate(basidia_count = readr::parse_number(as.character(basidia_count))) %>%
  filter(!is.na(basidia_count))
# Trying to pull section data into separate datasets

```


```{r}
library(readr)
library(dplyr)
library(ggplot2)

# Read your CSV
basidia <- read_csv("basidia_filled_rt.csv")

# Filter to specimens with BOTH avg_length and avg_width
basidia_clean <- basidia %>%
  filter(!is.na(avg_length), !is.na(avg_width))

# Define your color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Scatter plot
ggplot(basidia_clean, aes(x = avg_width, y = avg_length, color = sectionid)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Average Basidia Dimensions by Specimen",
    x = "Average Basidia Width (µm)",
    y = "Average Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14)

```

```{r}
#making a box and whisker plot of all the Vaginatae basidia lengths

library(dplyr)
library(ggplot2)
library(stringr)

# Start with your dataset
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to Vaginatae with complete basidia length data
vag <- basidia %>%
  filter(sectionid == "Vaginatae",
         !is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Create formatted species labels:
# Convert "Amanita fusca" → "A. fusca"
# Keep "Amanita sp aurorae" → "A. sp aurorae"
vag <- vag %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A."),
    species_short = factor(species_short, levels = unique(species_short))
  )

# Build the plot
ggplot(vag, aes(x = species_short)) +
  # Whisker line from min to max
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = species_short), linewidth = 1) +
  # Point for average
  geom_point(aes(y = avg_length), size = 3, color = "black") +
  labs(
    title = "Basidia Length Range per Specimen (Vaginatae Only)",
    x = "Species",
    y = "Basidia Length (µm)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length data
all_clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Create species labels: "Amanita fusca" → "A. fusca"
all_clean <- all_clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort by section, then species, then specimen_id
all_clean <- all_clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(all_clean, aes(x = specimen_id, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = specimen_id), linewidth = 1) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  scale_x_discrete(labels = all_clean$species_short) +
  labs(
    title = "Basidia Length Range per Specimen (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )
```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length data
all_clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Format species names: "Amanita fusca" → "A. fusca"
all_clean <- all_clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Reorder species by section first, then species name
all_clean <- all_clean %>%
  arrange(sectionid, species_short) %>%
  mutate(species_short = factor(species_short, levels = unique(species_short)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(all_clean, aes(x = species_short, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = species_short), linewidth = 1) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Basidia Length Range per Specimen (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

```

```{r}
#the average of the average within each section, the average minimum, the average maximum
library(dplyr)
library(ggplot2)
library(readr)

basidia <- read_csv("basidia_filled_mh.csv")

# Filter to rows with complete basidia length data
clean <- basidia %>%
  filter(!is.na(min_length),
         !is.na(avg_length),
         !is.na(max_length))

# Summarize by section
section_summary <- clean %>%
  group_by(sectionid) %>%
  summarise(
    mean_min = mean(min_length, na.rm = TRUE),
    mean_avg = mean(avg_length, na.rm = TRUE),
    mean_max = mean(max_length, na.rm = TRUE)
  ) %>%
  arrange(sectionid)

# Color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Plot
ggplot(section_summary, aes(x = sectionid, color = sectionid)) +
  # Whisker from mean min to mean max
  geom_segment(aes(y = mean_min, yend = mean_max,
                   xend = sectionid), linewidth = 1.2) +
  # Point for mean average
  geom_point(aes(y = mean_avg), size = 4) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "Average Basidia Length Range per Amanita Section",
    x = "Section",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)
install.packages("patchwork")
library(patchwork)   # for combining plots

# Read data
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with complete basidia length & width
clean <- basidia %>%
  filter(!is.na(avg_length),
         !is.na(avg_width),
         !is.na(min_length),
         !is.na(max_length))

# Format species names: "Amanita fusca" → "A. fusca"
clean <- clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort specimens by section, then species, then specimen_id
clean <- clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# -----------------------------
# Panel A: Scatterplot
# -----------------------------
p1 <- ggplot(clean, aes(x = avg_width, y = avg_length, color = sectionid)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = section_colors) +
  labs(
    title = "A. Scatterplot: Basidia Avg Width vs Avg Length",
    x = "Average Basidia Width (µm)",
    y = "Average Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14)

# -----------------------------
# Panel B: Specimen Whiskers
# -----------------------------
p2 <- ggplot(clean, aes(x = specimen_id, color = sectionid)) +
  geom_segment(aes(y = min_length, yend = max_length,
                   xend = specimen_id), linewidth = 1.2) +
  geom_point(aes(y = avg_length), size = 3) +
  scale_color_manual(values = section_colors) +
  scale_x_discrete(labels = clean$species_short) +
  labs(
    title = "B. Specimen Basidia Length Ranges (Sorted by Section)",
    x = "Species",
    y = "Basidia Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

# -----------------------------
# Combine Panels
# -----------------------------
p1 / p2   # stacked vertically

```

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(readr)

# Read data
basidia <- read_csv("basidia_filled_mh.csv")

# Filter to specimens with basidia_count
clean <- basidia %>%
  filter(!is.na(basidia_count))

# Create species labels: "Amanita fusca" → "A. fusca"
clean <- clean %>%
  mutate(
    species_short = str_replace(species, "^Amanita", "A.")
  )

# Sort by section, then species, then specimen_id
clean <- clean %>%
  arrange(sectionid, species_short, specimen_id) %>%
  mutate(specimen_id = factor(specimen_id, levels = unique(specimen_id)))

# Section color palette
section_colors <- c(
  "Roanokensis" = "magenta",
  "Phalloideae" = "blue",
  "Vaginatae"   = "yellow",
  "Amanita"     = "red",
  "Validae"     = "green",
  "Caesareae"   = "orange",
  "Amidella"    = "purple"
)

# Bar graph
ggplot(clean, aes(x = specimen_id, y = basidia_count, fill = sectionid)) +
  geom_col(color = "black") +
  scale_fill_manual(values = section_colors) +
  scale_x_discrete(labels = clean$species_short) +
  labs(
    title = "Number of Basidia Measured per Specimen",
    x = "Species",
    y = "Basidia Count",
    fill = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

```
Visualizing the sterigma count data

This first code chunk outputs a proportion of sterigma count per basidia which is cool but misleading for Caesareae
```{r}
## ============================================================
## Sterigma number variation across Amanita sections
## Reproducible visualization script
## This script visualizes proportion of sterigma count in each section which is interesting but is a bit inflated for sections like Caesareae where there isn't actually that many basidia that were counted
## ============================================================

## ---- setup ----
library(tidyverse)

## Set path to your CSV (edit as needed)
data_path <- "basidia_filled_mh.csv"

## ---- read data ----
basidia <- read_csv(data_path, show_col_types = FALSE)

## ---- select & clean sterigma data ----
## Keep only rows where basidia were actually counted
ster_cols <- c("4s/basidia", "3s/basidia", "2s/basidia", "s/basidia")

basidia_clean <- basidia %>%
  filter(if_any(all_of(ster_cols), ~ !is.na(.)))

## ---- pivot sterigma counts to long format ----
ster_long <- basidia_clean %>%
  select(specimen_id, species, sectionid, basidia_count, all_of(ster_cols)) %>%
  pivot_longer(
    cols = all_of(ster_cols),
    names_to = "sterigma_class",
    values_to = "count"
  ) %>%
  filter(!is.na(count), count > 0) %>%
  mutate(
    sterigmata = case_when(
      sterigma_class == "4s/basidia" ~ 4,
      sterigma_class == "3s/basidia" ~ 3,
      sterigma_class == "2s/basidia" ~ 2,
      sterigma_class == "s/basidia"  ~ 1
    )
  )

## ---- expand to per-basidium observations ----
## This allows true distributions instead of summarized counts
ster_expanded <- ster_long %>%
  uncount(weights = count)

## ============================================================
## FIGURE 1: Distribution of sterigma numbers by section
## ============================================================

ggplot(ster_expanded, aes(x = sectionid, y = sterigmata)) +
  geom_violin(
    fill = "grey85",
    color = "grey30",
    trim = FALSE
  ) +
  geom_jitter(
    width = 0.15,
    size = 0.5,
    alpha = 0.3
  ) +
  stat_summary(
    fun = median,
    geom = "point",
    size = 2,
    color = "black"
  ) +
  labs(
    x = "Amanita section",
    y = "Number of sterigmata per basidium",
    title = "Sterigma number variation across Amanita sections",
    subtitle = "Each point represents a single basidium"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

## ============================================================
## FIGURE 2: Proportional sterigma composition by section
## ============================================================

ster_props <- ster_long %>%
  group_by(sectionid, sterigmata) %>%
  summarise(total = sum(count), .groups = "drop") %>%
  group_by(sectionid) %>%
  mutate(prop = total / sum(total))

ggplot(ster_props, aes(x = sectionid, y = prop, fill = factor(sterigmata))) +
  geom_col(color = "black") +
  scale_fill_brewer(
    palette = "Set2",
    name = "Sterigmata\nper basidium"
  ) +
  labs(
    x = "Amanita section",
    y = "Proportion of basidia",
    title = "Section-level sterigma composition in Amanita"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

## ============================================================
## FIGURE 3: Sterigma diversity (developmental lability)
## ============================================================

ster_diversity <- ster_props %>%
  group_by(sectionid) %>%
  summarise(
    shannon_diversity = -sum(prop * log(prop))
  )

ggplot(ster_diversity, aes(x = sectionid, y = shannon_diversity)) +
  geom_col(fill = "grey70", color = "black") +
  labs(
    x = "Amanita section",
    y = "Sterigma diversity (Shannon index)",
    title = "Developmental diversity of sterigma number by section"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

## ============================================================
## END SCRIPT
## ============================================================
```

Making a section newick tree to use in visualizations
```{r}
library(ape)
library(nlme)

sectree <- read.tree('sec_tre.nwk')
plot(sectree)
```

Total basidia counted graph:
```{r}
library(tidyverse)
library(ggplot2)
library(ggtree)
library(patchwork)
library(tidyr)

basidia <- read_csv("basidia_filled_mh.csv", show_col_types = FALSE)

#Identify sterigma count columns
ster_cols <- c("4s/basidia", "3s/basidia", "2s/basidia", "s/basidia")

#Calculate total basidia observed for each specimen
basidia_totals <- basidia %>%
  mutate(
    total_basidia = rowSums(across(all_of(ster_cols)), na.rm = TRUE)
  ) %>%
  filter(total_basidia > 0)

#Sum total basidia observed per section
section_totals <- basidia_totals %>%
  group_by(sectionid) %>%
  summarise(
    total_basidia = sum(total_basidia),
    n_specimens = n(),
    .groups = "drop"
  )

#Add all sections
tree_sections <- sectree$tip.label
section_totals_full <- section_totals %>%
  complete(
    sectionid = tree_sections,
    fill = list(total_basidia = 0)
  )

#Oragnize sections phylogenetically
section_totals_full <- section_totals_full %>%
  mutate(sectionid = factor(sectionid, levels = sectree$tip.label))

#Plot section totals

p_bar <- ggplot(section_totals_full,
                aes(x = sectionid, y = total_basidia, fill = sectionid)) +
  geom_col(color = "black") +
  geom_text(
    aes(label = ifelse(total_basidia > 0, total_basidia, "")),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Total basidia observed"
  ) +
  theme_classic() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

#Plot stacked bar plot
ggplot(
  basidia_totals,
  aes(x = sectionid, y = total_basidia, fill = specimen_id)
) +
  geom_col(color = "black", linewidth = 0.1) +
  labs(
    x = "Amanita section",
    y = "Total number of basidia observed",
    fill = "Specimen ID",
    title = "Total basidia per section, colored by contributing specimen"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```


