title: "micro_analyze"
output: html_document
date: "2026-02-26"

All of this needs labeling and descriptions of what the code is doing:
```{r}

setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore") #set your working directory
setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore/basidiospore_by_specimen")
getwd() #use to see what your current working directory is
#rename each csv to spore30181, spore30182, etc for easier coding
spores30181 <- read.csv("30181_basidiospore.csv")
spores30182 <- read.csv("30182_basidiospore.csv")
spores30183 <- read.csv("30183_basidiospore.csv")
spores30184 <- read.csv("30184_basidiospore.csv")
spores30185 <- read.csv("30185_basidiospore.csv")
spores30186 <- read.csv("30186_basidiospore.csv")
spores30187 <- read.csv("30187_basidiospore.csv")
spores30188 <- read.csv("30188_basidiospore.csv")
spores30189 <- read.csv("30189_basidiospore.csv")
spores30190 <- read.csv("30190_basidiospore.csv")
spores30191 <- read.csv("30191_basidiospore.csv")
spores30192 <- read.csv("30192_basidiospore.csv")
spores30193 <- read.csv("30193_basidiospore.csv")
spores30194 <- read.csv("30194_basidiospore.csv")
spores30195 <- read.csv("30195_basidiospore.csv")
spores30196 <- read.csv("30196_basidiospore.csv")
spores30197 <- read.csv("30197_basidiospore.csv")
spores30198 <- read.csv("30198_basidiospore.csv")
spores30199 <- read.csv("30199_basidiospore.csv")
spores30201 <- read.csv("30201_basidiospore.csv")
spores30202 <- read.csv("30202_basidiospore.csv")
spores30203 <- read.csv("30203_basidiospore.csv")
spores30204 <- read.csv("30204_basidiospore.csv")
spores30205 <- read.csv("30205_basidiospore.csv")
spores30206 <- read.csv("30206_basidiospore.csv")
spores30207 <- read.csv("30207_basidiospore.csv")
spores30208 <- read.csv("30208_basidiospore.csv")
spores30209 <- read.csv("30209_basidiospore.csv")
spores30212 <- read.csv("30212_basidiospore.csv")
spores30213 <- read.csv("30213_basidiospore.csv")
spores30214 <- read.csv("30214_basidiospore.csv")
spores30215 <- read.csv("30215_basidiospore.csv")
spores30221 <- read.csv("30221_basidiospore.csv")
spores30222 <- read.csv("30222_basidiospore.csv")
spores30226 <- read.csv("30226_basidiospore.csv")
spores30230 <- read.csv("30230_basidiospore.csv")
spores30238 <- read.csv("30238_basidiospore.csv")
spores30239 <- read.csv("30239_basidiospore.csv")
spores30244 <- read.csv("30244_basidiospore.csv")
spores30249 <- read.csv("30249_basidiospore.csv")
spores30300 <- read.csv("30300_basidiospore.csv")
spores30295 <- read.csv("30295_basidiospore.csv")
spores30276 <- read.csv("30276_basidiospore.csv")
spores30274 <- read.csv("30274_basidiospore.csv")
spores30248 <- read.csv("30248_basidiospore.csv")
spores30245 <- read.csv("30245_basidiospore.csv")


# $Sample to add new column with speices ID to each basidiospore measurement
spores30181$Sample <- "30181"
spores30182$Sample <- "30182"
spores30183$Sample <- "30183"
spores30184$Sample <- "30184"
spores30185$Sample <- "30185"
spores30186$Sample <- "30186"
spores30187$Sample <- "30187"
spores30188$Sample <- "30188"
spores30189$Sample <- "30189"
spores30190$Sample <- "30190"
spores30191$Sample <- "30191"
spores30192$Sample <- "30192"
spores30193$Sample <- "30193"
spores30194$Sample <- "30194"
spores30195$Sample <- "30195"
spores30196$Sample <- "30196"
spores30197$Sample <- "30197"
spores30198$Sample <- "30198"
spores30199$Sample <- "30199"
spores30201$Sample <- "30201"
spores30202$Sample <- "30202"
spores30203$Sample <- "30203"
spores30204$Sample <- "30204"
spores30205$Sample <- "30205"
spores30206$Sample <- "30206"
spores30207$Sample <- "30207"
spores30208$Sample <- "30208"
spores30209$Sample <- "30209"
spores30212$Sample <- "30212"
spores30213$Sample <- "30213"
spores30214$Sample <- "30214"
spores30215$Sample <- "30215"
spores30221$Sample <- "30221"
spores30222$Sample <- "30222"
spores30226$Sample <- "30226"
spores30230$Sample <- "30230"
spores30238$Sample <- "30238"
spores30239$Sample <- "30239"
spores30244$Sample <- "30244"
spores30249$Sample <- "30249"
spores30300$Sample <- "30300"
spores30295$Sample <- "30295"
spores30276$Sample <- "30276"
spores30274$Sample <- "30274"
spores30248$Sample <- "30248"
spores30245$Sample <- "30245"

str(spores30181) #sanity check to make sure data loaded

head(spores30181) #sanity check to make sure data loaded

#combines each specimen to one graph
all_spores <- rbind(spores30181, spores30182, spores30183, spores30184, spores30185, spores30186, spores30187, spores30188, spores30189, spores30190, spores30191, spores30192, spores30193, spores30194, spores30195, spores30196, spores30197, spores30198, spores30199, spores30201, spores30202, spores30203, spores30204, spores30205, spores30206, spores30207, spores30208, spores30209, spores30212, spores30213, spores30214, spores30215, spores30221, spores30222, spores30226, spores30230, spores30238, spores30239, spores30244, spores30249, spores30300, spores30295, spores30276, spores30274, spores30248, spores30245)
library(ggplot2)

#groups each specimen by section
group_map <- c(
  "30181"="Phalloideae", "30182"="Phalloideae", "30183"="Roanokensis",
  "30184"="Phalloideae", "30185"="Vaginatae", "30186"="Vaginatae",
  "30187"="Phalloideae", "30188"="Vaginatae", "30189"="Vaginatae",
  "30190"="Amanita", "30191"="Amanita", "30192"="Amanita",
  "30193"="Vaginatae", "30194"="Vaginatae", "30195"="Phalloideae",
  "30196"="Amanita", "30197"="Amanita", "30198"="Vaginatae",
  "30199"="Phalloideae", "30201"="Validae", "30202"="Validae",
  "30203"="Phalloideae", "30204"="Phalloideae", "30205"="Validae",
  "30206"="Validae", "30207"="Validae", "30208"="Validae",
  "30209"="Phalloideae", "30212"="Amidella", "30213"="Validae",
  "30214"="Validae", "30215"="Vaginatae",
  "30221"="Amanita", "30222"="Amanita", "30226"="Amanita",
  "30230"="Validae", "30238"="Validae", "30239"="Caesareae",
  "30244"="Vaginatae", "30249"="Amanita", "30300"="Amidella", "30295"="Caesareae", "30276"="Phalloideae", "30274"="Amidella", "30248"="Validae", "30245"="Vaginatae"
)
all_spores$Group <- group_map[all_spores$Sample]

# Set custom legend order
all_spores$Group <- factor(
  all_spores$Group,
  levels = c(
    "Roanokensis",
    "Amidella",
    "Validae",
    "Phalloideae",
    "Amanita",
    "Vaginatae",
    "Caesareae"
  )
)


plot <- ggplot(all_spores, aes(x = Width, y = Length, fill = Group)) +
  geom_jitter(shape = 21, size = 2, alpha = 0.7, width = 0.25, height = 0.25) +

  #formating
  scale_fill_manual(
    name = "Section",
    values = c(
      "Roanokensis"="magenta",
      "Phalloideae"="blue",
      "Vaginatae"="yellow",
      "Amanita"="red",
      "Validae"="green",
      "Caesareae"="orange",
      "Amidella"="purple"
    ),
      aesthetics = "fill"
  ) +
  guides(color = "none") +
  xlim(0, 15) +
  ylim(0, 20) +
  theme_minimal() +
  labs(
    title = "Basidiospore Measurements",
    x = "Width (µm)",
    y = "Length (µm)"
  )

plot (plot)

#ggsave("basidiospore_measurementes_20.png", plot, width = 10, height = 7, dpi = 300) to get .png file
```

```{r}

setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore") #set your working directory
setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore/basidiospore_by_specimen")
getwd() #use to see what your current working directory is
#rename each csv to spore30181, etc for easier coding
spores30181 <- read.csv("30181_basidiospore.csv")
spores30182 <- read.csv("30182_basidiospore.csv")
spores30183 <- read.csv("30183_basidiospore.csv")
spores30184 <- read.csv("30184_basidiospore.csv")
spores30185 <- read.csv("30185_basidiospore.csv")
spores30186 <- read.csv("30186_basidiospore.csv")
spores30187 <- read.csv("30187_basidiospore.csv")
spores30188 <- read.csv("30188_basidiospore.csv")
spores30189 <- read.csv("30189_basidiospore.csv")
spores30190 <- read.csv("30190_basidiospore.csv")
spores30191 <- read.csv("30191_basidiospore.csv")
spores30192 <- read.csv("30192_basidiospore.csv")
spores30193 <- read.csv("30193_basidiospore.csv")
spores30194 <- read.csv("30194_basidiospore.csv")
spores30195 <- read.csv("30195_basidiospore.csv")
spores30196 <- read.csv("30196_basidiospore.csv")
spores30197 <- read.csv("30197_basidiospore.csv")
spores30198 <- read.csv("30198_basidiospore.csv")
spores30199 <- read.csv("30199_basidiospore.csv")
spores30201 <- read.csv("30201_basidiospore.csv")
spores30202 <- read.csv("30202_basidiospore.csv")
spores30203 <- read.csv("30203_basidiospore.csv")
spores30204 <- read.csv("30204_basidiospore.csv")
spores30205 <- read.csv("30205_basidiospore.csv")
spores30206 <- read.csv("30206_basidiospore.csv")
spores30207 <- read.csv("30207_basidiospore.csv")
spores30208 <- read.csv("30208_basidiospore.csv")
spores30209 <- read.csv("30209_basidiospore.csv")
spores30212 <- read.csv("30212_basidiospore.csv")
spores30213 <- read.csv("30213_basidiospore.csv")
spores30214 <- read.csv("30214_basidiospore.csv")
spores30215 <- read.csv("30215_basidiospore.csv")
spores30221 <- read.csv("30221_basidiospore.csv")
spores30222 <- read.csv("30222_basidiospore.csv")
spores30226 <- read.csv("30226_basidiospore.csv")
spores30230 <- read.csv("30230_basidiospore.csv")
spores30238 <- read.csv("30238_basidiospore.csv")
spores30239 <- read.csv("30239_basidiospore.csv")
spores30244 <- read.csv("30244_basidiospore.csv")
spores30249 <- read.csv("30249_basidiospore.csv")
spores30300 <- read.csv("30300_basidiospore.csv")
spores30295 <- read.csv("30295_basidiospore.csv")
spores30276 <- read.csv("30276_basidiospore.csv")
spores30274 <- read.csv("30274_basidiospore.csv")
spores30248 <- read.csv("30248_basidiospore.csv")
spores30245 <- read.csv("30245_basidiospore.csv")


# $Sample to add new column with speices ID to each basidiospore measurement
spores30181$Sample <- "30181"
spores30182$Sample <- "30182"
spores30183$Sample <- "30183"
spores30184$Sample <- "30184"
spores30185$Sample <- "30185"
spores30186$Sample <- "30186"
spores30187$Sample <- "30187"
spores30188$Sample <- "30188"
spores30189$Sample <- "30189"
spores30190$Sample <- "30190"
spores30191$Sample <- "30191"
spores30192$Sample <- "30192"
spores30193$Sample <- "30193"
spores30194$Sample <- "30194"
spores30195$Sample <- "30195"
spores30196$Sample <- "30196"
spores30197$Sample <- "30197"
spores30198$Sample <- "30198"
spores30199$Sample <- "30199"
spores30201$Sample <- "30201"
spores30202$Sample <- "30202"
spores30203$Sample <- "30203"
spores30204$Sample <- "30204"
spores30205$Sample <- "30205"
spores30206$Sample <- "30206"
spores30207$Sample <- "30207"
spores30208$Sample <- "30208"
spores30209$Sample <- "30209"
spores30212$Sample <- "30212"
spores30213$Sample <- "30213"
spores30214$Sample <- "30214"
spores30215$Sample <- "30215"
spores30221$Sample <- "30221"
spores30222$Sample <- "30222"
spores30226$Sample <- "30226"
spores30230$Sample <- "30230"
spores30238$Sample <- "30238"
spores30239$Sample <- "30239"
spores30244$Sample <- "30244"
spores30249$Sample <- "30249"
spores30300$Sample <- "30300"
spores30295$Sample <- "30295"
spores30276$Sample <- "30276"
spores30274$Sample <- "30274"
spores30248$Sample <- "30248"
spores30245$Sample <- "30245"

str(spores30181) #sanity check to make sure data loaded

head(spores30181) #sanity check to make sure data loaded

#combines each specimen to one graph
all_spores <- rbind(spores30181, spores30182, spores30183, spores30184, spores30185, spores30186, spores30187, spores30188, spores30189, spores30190, spores30191, spores30192, spores30193, spores30194, spores30195, spores30196, spores30197, spores30198, spores30199, spores30201, spores30202, spores30203, spores30204, spores30205, spores30206, spores30207, spores30208, spores30209, spores30212, spores30213, spores30214, spores30215, spores30221, spores30222, spores30226, spores30230, spores30238, spores30239, spores30244, spores30249, spores30300, spores30295, spores30276, spores30274, spores30248, spores30245)
library(ggplot2)

#groups each specimen by section
group_map <- c(
  "30181"="Phalloideae", "30182"="Phalloideae", "30183"="Roanokensis",
  "30184"="Phalloideae", "30185"="Vaginatae", "30186"="Vaginatae",
  "30187"="Phalloideae", "30188"="Vaginatae", "30189"="Vaginatae",
  "30190"="Amanita", "30191"="Amanita", "30192"="Amanita",
  "30193"="Vaginatae", "30194"="Vaginatae", "30195"="Phalloideae",
  "30196"="Amanita", "30197"="Amanita", "30198"="Vaginatae",
  "30199"="Phalloideae", "30201"="Validae", "30202"="Validae",
  "30203"="Phalloideae", "30204"="Phalloideae", "30205"="Validae",
  "30206"="Validae", "30207"="Validae", "30208"="Validae",
  "30209"="Phalloideae", "30212"="Amidella", "30213"="Validae",
  "30214"="Validae", "30215"="Vaginatae",
  "30221"="Amanita", "30222"="Amanita", "30226"="Amanita",
  "30230"="Validae", "30238"="Validae", "30239"="Caesareae",
  "30244"="Vaginatae", "30249"="Amanita", "30300"="Amidella", "30295"="Caesareae", "30276"="Phalloideae", "30274"="Amidella", "30248"="Validae", "30245"="Vaginatae"
)
all_spores$Group <- group_map[all_spores$Sample]

# Set custom legend order
all_spores$Group <- factor(
  all_spores$Group,
  levels = c(
    "Roanokensis",
    "Amidella",
    "Validae",
    "Phalloideae",
    "Amanita",
    "Vaginatae",
    "Caesareae"
  )
)


plot <- ggplot(all_spores, aes(x = Width, y = Length, fill = Group)) +
  geom_jitter(shape = 21, size = 2, alpha = 0.7, width = 0.25, height = 0.25) +

  #formating
  scale_fill_manual(
    name = "Section",
    values = c(
      "Roanokensis"="magenta",
      "Phalloideae"="blue",
      "Vaginatae"="yellow",
      "Amanita"="red",
      "Validae"="green",
      "Caesareae"="orange",
      "Amidella"="purple"
    ),
      aesthetics = "fill"
  ) +
  guides(color = "none") +
  xlim(5, 15) +
  ylim(5, 20) +
  theme_minimal() +
  labs(
    title = "Basidiospore Measurements",
    x = "Width (µm)",
    y = "Length (µm)"
  )

plot (plot)

#ggsave("basidiospore_measurementes_20.png", plot, width = 10, height = 7, dpi = 300) to get .png file
ggsave("basidiospore_measurementes_2.png", plot, width = 10, height = 7, dpi = 300)
```

```{r}
library(dplyr)
library(ggplot2)

# Count number of spores per section
section_counts <- all_spores %>%
  group_by(Group) %>%
  summarise(n_spores = n())

# (Optional) sanity check
table(all_spores$Group, useNA = "ifany")

# Set custom legend + x‑axis order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

section_counts$Group <- factor(section_counts$Group, levels = section_order)

# Bar plot
bar_plot <- ggplot(section_counts, aes(x = Group, y = n_spores, fill = Group)) +
  geom_col(color = "black", width = 0.7) +
  scale_fill_manual(
    values = c(
      "Roanokensis"="magenta",
      "Phalloideae"="blue",
      "Vaginatae"="yellow",
      "Amanita"="red",
      "Validae"="green",
      "Caesareae"="orange",
      "Amidella"="purple"
    )
  ) +
  labs(
    title = "Number of Basidiospore Measurements per Section",
    x = "Section",
    y = "Number of Measurements"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    
  )

bar_plot

#ggsave("figures/basidiospore_spores_per_section.png", bar_plot, width = 10, height = 7, dpi = 300)


```
```{r}
library(dplyr)
library(ggplot2)

# Count unique species per section
species_per_section <- all_spores2 %>%
  group_by(sectionid) %>%
  summarise(
    n_species = n_distinct(Species_clean)
  )

# Apply your custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

species_per_section$sectionid <- factor(
  species_per_section$sectionid,
  levels = section_order
)

# Section colors
section_colors <- c(
  "Roanokensis"="magenta",
  "Phalloideae"="blue",
  "Vaginatae"="yellow",
  "Amanita"="red",
  "Validae"="green",
  "Caesareae"="orange",
  "Amidella"="purple"
)

# Plot
species_plot <- ggplot(species_per_section,
                       aes(x = sectionid, y = n_species, fill = sectionid)) +
  geom_col(color = "black", width = 0.7) +
  scale_fill_manual(values = section_colors) +
  scale_y_continuous(breaks = 0:max(species_per_section$n_species)) +
  labs(
    title = "Number of Unique Species per Section",
    x = "Section",
    y = "Number of Unique Species"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    legend.position = "none"
  )

species_plot


```

```{r}
library(dplyr)
library(ggplot2)

# ------------------------------------------------------------
# 1. Count the number of UNIQUE specimens per section
#    - all_spores must already contain:
#         * Sample  (specimen ID)
#         * Group   (section assignment)
# ------------------------------------------------------------

samples_per_section <- all_spores %>%
  group_by(Group) %>%
  summarise(n_samples = n_distinct(Sample))

# View the table to confirm whole-number counts
samples_per_section


# ------------------------------------------------------------
# 2. Plot: Number of specimens per section
#    - Uses your exact color scheme
#    - Forces y-axis to use whole-number ticks only
# ------------------------------------------------------------

sample_plot <- ggplot(samples_per_section, aes(x = Group, y = n_samples, fill = Group)) +
  
  # Draw bars
  geom_col(color = "black", width = 0.7) +
  
  # Manual color palette (same as your other plots)
  scale_fill_manual(
    values = c(
      "Roanokensis" = "magenta",
      "Phalloideae"  = "blue",
      "Vaginatae"   = "yellow",
      "Amanita"     = "red",
      "Validae"     = "green",
      "Caesareae"   = "orange",
      "Amidella"    = "purple"
    )
  ) +
  
  # Force y-axis to show ONLY whole numbers
  scale_y_continuous(breaks = 0:max(samples_per_section$n_samples)) +
  
  # Labels
  labs(
    title = "Number of Specimens per Section",
    x = "Section",
    y = "Number of Specimens"
  ) +
  
  # Theme adjustments
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
  )

# Custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

# Apply ordering to the summary table
samples_per_section$Group <- factor(samples_per_section$Group,
                                    levels = section_order)

# Display the plot
sample_plot


#ggsave("figures/basidiospore_specimens_per_section.png", sample_plot, width = 10, height = 7, dpi = 300)


```
```{r}
# Rebuild Group from group_map using clean character Sample IDs
all_spores$Sample <- as.character(all_spores$Sample)
group_map <- c(
  "30181"="Phalloideae", "30182"="Phalloideae", "30183"="Roanokensis",
  "30184"="Phalloideae", "30185"="Vaginatae", "30186"="Vaginatae",
  "30187"="Phalloideae", "30188"="Vaginatae", "30189"="Vaginatae",
  "30190"="Amanita", "30191"="Amanita", "30192"="Amanita",
  "30193"="Vaginatae", "30194"="Vaginatae", "30195"="Phalloideae",
  "30196"="Amanita", "30197"="Amanita", "30198"="Vaginatae",
  "30199"="Phalloideae", "30201"="Validae", "30202"="Validae",
  "30203"="Phalloideae", "30204"="Phalloideae", "30205"="Validae",
  "30206"="Validae", "30207"="Validae", "30208"="Validae",
  "30209"="Phalloideae", "30212"="Amidella", "30213"="Validae",
  "30214"="Validae", "30215"="Vaginatae",
  "30221"="Amanita", "30222"="Amanita", "30226"="Amanita",
  "30230"="Validae", "30238"="Validae", "30239"="Caesareae",
  "30244"="Vaginatae", "30249"="Amanita",
  "30300"="Amidella", "30295"="Caesareae",
  "30276"="Phalloideae", "30274"="Amidella",
  "30248"="Validae", "30245"="Vaginatae"
)

all_spores$Group <- group_map[all_spores$Sample]

# Trim any stray whitespace, just in case
all_spores$Group <- trimws(all_spores$Group)

# Check what actually exists
unique(all_spores$Group)
table(all_spores$Group, useNA = "ifany")

```

```{r}
library(dplyr)
library(ggplot2)
setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
knitr::opts_knit$set(root.dir = "~/Desktop/wiscam_microscopy/data/rt_basidiospore")

# 1. Load species lookup table
species_lookup <- read.csv("basidiospore.csv", stringsAsFactors = FALSE)

# 2. Rename specimen_id to Sample so it matches all_spores
species_lookup <- species_lookup %>%
  rename(Sample = specimen_id)

# 3. Ensure Sample is character
species_lookup$Sample <- as.character(species_lookup$Sample)

# 4. Merge species names into the main dataset
all_spores2 <- all_spores %>%
  left_join(species_lookup, by = "Sample")

# Reapply Group after join to fix NA values
all_spores2$Group <- group_map[all_spores2$Sample]

# Trim whitespace just in case
all_spores2$Group <- trimws(all_spores2$Group)

# Check results
unique(all_spores2$Group)
table(all_spores2$Group, useNA = "ifany")

# 5. Clean species names into "A. species" format
all_spores2$Species_clean <- sub("^Amanita ", "A. ", all_spores2$species)

# 6. Count spores per species (combining repeated specimens)
species_counts <- all_spores2 %>%
  group_by(Species_clean, Group) %>%
  summarise(n_spores = n(), .groups = "drop")

# 7. Sort species by section (Group), then alphabetically within section
species_counts$Species_clean <- factor(
  species_counts$Species_clean,
  levels = species_counts %>%
    arrange(Group, Species_clean) %>%
    pull(Species_clean)
)

# Custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

# Apply ordering to Group
species_counts$Group <- factor(species_counts$Group, levels = section_order)

# Reorder species by section, then alphabetically
species_counts$Species_clean <- factor(
  species_counts$Species_clean,
  levels = species_counts %>%
    arrange(Group, Species_clean) %>%
    pull(Species_clean)
)

# 8. Bar plot by species, color-coded by section
species_plot <- ggplot(species_counts,
                       aes(x = Species_clean,
                           y = n_spores,
                           fill = Group)) +
  geom_col(color = "black") +
  scale_fill_manual(
    values = c(
      "Roanokensis"="magenta",
      "Phalloideae"="blue",
      "Vaginatae"="yellow",
      "Amanita"="red",
      "Validae"="green",
      "Caesareae"="orange",
      "Amidella"="purple"
    )
  ) +
  labs(
    title = "Total Basidiospore Measurements per Species",
    x = "Species",
    y = "Number of Measurements"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

species_plot

# 9. Save the plot — absolute path avoids R Markdown directory issues
ggsave(
  filename = "basidiospore_species_barplot.png",
  plot = species_plot,
  width = 12,
  height = 8,
  dpi = 300,
  path = path.expand("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
)

```
#add, solaniolens,

```{r}
library(dplyr)
library(ggplot2)
library(stringr)


# ------------------------------------------------------------
# 1. Load all basidiospore CSVs automatically (correct version)
# ------------------------------------------------------------

spore_files <- list.files(
  "data/rt_basidiospore/basidiospore_by_specimen",
  pattern = "*.csv",
  full.names = TRUE
)

# Read each file AND add Sample ID inside the loop
all_spores <- lapply(spore_files, function(f) {
  df <- read.csv(f)
  df$Sample <- str_extract(basename(f), "^\\d+")
  df
}) %>% bind_rows()

# ------------------------------------------------------------
# 2. Load species + section lookup table
# ------------------------------------------------------------

lookup <- read.csv("data/rt_basidiospore/basidiospore.csv",
                   stringsAsFactors = FALSE)

# Ensure Sample is character for joining
lookup$Sample <- as.character(lookup$specimen_id)
unique(lookup$sectionid)

# ------------------------------------------------------------
# 3. Merge lookup into spore dataset
# ------------------------------------------------------------

all_spores2 <- all_spores %>%
  left_join(lookup, by = "Sample")

# ------------------------------------------------------------
# 4. Clean species names (Amanita → A.)
# ------------------------------------------------------------

all_spores2$Species_clean <- str_replace(all_spores2$species, "^Amanita ", "A. ")

# ------------------------------------------------------------
# 5. Remove any rows missing section assignment
# ------------------------------------------------------------

all_spores2 <- all_spores2 %>% filter(!is.na(sectionid))

# ------------------------------------------------------------
# 6. Normalize + order section names
# ------------------------------------------------------------

# Custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

all_spores2$sectionid <- factor(all_spores2$sectionid,
                                levels = section_order)

# ------------------------------------------------------------
# 7. Order species by section, then alphabetically
# ------------------------------------------------------------

all_spores2$Species_clean <- factor(
  all_spores2$Species_clean,
  levels = all_spores2 %>%
    distinct(Species_clean, sectionid) %>%
    arrange(sectionid, Species_clean) %>%
    pull(Species_clean)
)

# ------------------------------------------------------------
# 8. Define section colors
# ------------------------------------------------------------

section_colors <- c(
  "Roanokensis"="magenta",
  "Phalloideae"="blue",
  "Vaginatae"="yellow",
  "Amanita"="red",
  "Validae"="green",
  "Caesareae"="orange",
  "Amidella"="purple"
)

# ------------------------------------------------------------
# 9. Species-level boxplot (no mean point)
# ------------------------------------------------------------

ggplot(all_spores2, aes(x = Species_clean, y = Length, fill = sectionid)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, lwd = 0.4) +
  scale_fill_manual(values = section_colors) +
  labs(
    title = "Basidiospore Length Distribution by Species",
    x = "Species",
    y = "Basidiospore Length (µm)",
    fill = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

#ggsave
ggsave(
  filename = "basidiospore_species_boxplot.png",
  plot = last_plot(),
  width = 12,
  height = 8,
  dpi = 300,
  path = path.expand("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
)

```

```{r}
#same thing but width
library(dplyr)
library(ggplot2)
library(stringr)


# ------------------------------------------------------------
# 1. Load all basidiospore CSVs automatically (correct version)
# ------------------------------------------------------------

spore_files <- list.files(
  "data/rt_basidiospore/basidiospore_by_specimen",
  pattern = "*.csv",
  full.names = TRUE
)

# Read each file AND add Sample ID inside the loop
all_spores <- lapply(spore_files, function(f) {
  df <- read.csv(f)
  df$Sample <- str_extract(basename(f), "^\\d+")
  df
}) %>% bind_rows()

# ------------------------------------------------------------
# 2. Load species + section lookup table
# ------------------------------------------------------------

lookup <- read.csv("data/rt_basidiospore/basidiospore.csv",
                   stringsAsFactors = FALSE)

# Ensure Sample is character for joining
lookup$Sample <- as.character(lookup$specimen_id)
unique(lookup$sectionid)

# ------------------------------------------------------------
# 3. Merge lookup into spore dataset
# ------------------------------------------------------------

all_spores2 <- all_spores %>%
  left_join(lookup, by = "Sample")

# ------------------------------------------------------------
# 4. Clean species names (Amanita → A.)
# ------------------------------------------------------------

all_spores2$Species_clean <- str_replace(all_spores2$species, "^Amanita ", "A. ")

# ------------------------------------------------------------
# 5. Remove any rows missing section assignment
# ------------------------------------------------------------

all_spores2 <- all_spores2 %>% filter(!is.na(sectionid))

# ------------------------------------------------------------
# 6. Normalize + order section names
# ------------------------------------------------------------

# Custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

all_spores2$sectionid <- factor(all_spores2$sectionid,
                                levels = section_order)

# ------------------------------------------------------------
# 7. Order species by section, then alphabetically
# ------------------------------------------------------------

all_spores2$Species_clean <- factor(
  all_spores2$Species_clean,
  levels = all_spores2 %>%
    distinct(Species_clean, sectionid) %>%
    arrange(sectionid, Species_clean) %>%
    pull(Species_clean)
)

# ------------------------------------------------------------
# 8. Define section colors
# ------------------------------------------------------------

section_colors <- c(
  "Roanokensis"="magenta",
  "Phalloideae"="blue",
  "Vaginatae"="yellow",
  "Amanita"="red",
  "Validae"="green",
  "Caesareae"="orange",
  "Amidella"="purple"
)

# ------------------------------------------------------------
# 9. Species-level boxplot (no mean point)
# ------------------------------------------------------------

ggplot(all_spores2, aes(x = Species_clean, y = Width, fill = sectionid)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, lwd = 0.4) +
  scale_fill_manual(values = section_colors) +
  labs(
    title = "Basidiospore Width Distribution by Species",
    x = "Species",
    y = "Basidiospore Width (µm)",
    fill = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

#ggsave
ggsave(
  filename = "basidiospore_species_boxplot_2.png",
  plot = last_plot(),
  width = 12,
  height = 8,
  dpi = 300,
  path = path.expand("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
)

```

```{r}
#same thing but Q values

library(dplyr)
library(ggplot2)
library(stringr)


# ------------------------------------------------------------
# 1. Load all basidiospore CSVs automatically (correct version)
# ------------------------------------------------------------

spore_files <- list.files(
  "data/rt_basidiospore/basidiospore_by_specimen",
  pattern = "*.csv",
  full.names = TRUE
)

# Read each file AND add Sample ID inside the loop
all_spores <- lapply(spore_files, function(f) {
  df <- read.csv(f)
  df$Sample <- str_extract(basename(f), "^\\d+")
  df
}) %>% bind_rows()

# ------------------------------------------------------------
# 2. Load species + section lookup table
# ------------------------------------------------------------

lookup <- read.csv("data/rt_basidiospore/basidiospore.csv",
                   stringsAsFactors = FALSE)

# Ensure Sample is character for joining
lookup$Sample <- as.character(lookup$specimen_id)
unique(lookup$sectionid)

# ------------------------------------------------------------
# 3. Merge lookup into spore dataset
# ------------------------------------------------------------

all_spores2 <- all_spores %>%
  left_join(lookup, by = "Sample")

# ------------------------------------------------------------
# 4. Clean species names (Amanita → A.)
# ------------------------------------------------------------

all_spores2$Species_clean <- str_replace(all_spores2$species, "^Amanita ", "A. ")

# ------------------------------------------------------------
# 5. Remove any rows missing section assignment
# ------------------------------------------------------------

all_spores2 <- all_spores2 %>% filter(!is.na(sectionid))

# ------------------------------------------------------------
# 6. Normalize + order section names
# ------------------------------------------------------------

# Custom section order
section_order <- c(
  "Roanokensis",
  "Amidella",
  "Validae",
  "Phalloideae",
  "Amanita",
  "Vaginatae",
  "Caesareae"
)

all_spores2$sectionid <- factor(all_spores2$sectionid,
                                levels = section_order)

# ------------------------------------------------------------
# 7. Order species by section, then alphabetically
# ------------------------------------------------------------

all_spores2$Species_clean <- factor(
  all_spores2$Species_clean,
  levels = all_spores2 %>%
    distinct(Species_clean, sectionid) %>%
    arrange(sectionid, Species_clean) %>%
    pull(Species_clean)
)

# ------------------------------------------------------------
# 8. Define section colors
# ------------------------------------------------------------

section_colors <- c(
  "Roanokensis"="magenta",
  "Phalloideae"="blue",
  "Vaginatae"="yellow",
  "Amanita"="red",
  "Validae"="green",
  "Caesareae"="orange",
  "Amidella"="purple"
)

# ------------------------------------------------------------
# 9. Species-level boxplot (no mean point)
# ------------------------------------------------------------

ggplot(all_spores2, aes(x = Species_clean, y = Q..length.width., fill = sectionid)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, lwd = 0.4) +
  scale_fill_manual(values = section_colors) +
  labs(
    title = "Basidiospore Q-value Distribution by Species",
    x = "Species",
    y = "Basidiospore Q-value",
    fill = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic")
  )

#ggsave
ggsave(
  filename = "basidiospore_species_boxplot_3.png",
  plot = last_plot(),
  width = 12,
  height = 8,
  dpi = 300,
  path = path.expand("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
)

```


```{r}
ggplot(all_spores2, aes(x = Width, y = Length, color = sectionid)) +
  geom_jitter(width = 0.15, height = 0.15, alpha = 0.6, size = 1.8) +  # jittered points
  geom_abline(intercept = 0, slope = 1,
              color = "black", linetype = "dashed", linewidth = 0.7) + # Q = 1 line
  scale_color_manual(values = section_colors) +
  coord_cartesian(xlim = c(0, max(all_spores2$Width, na.rm = TRUE)),
                  ylim = c(0, max(all_spores2$Length, na.rm = TRUE))) +
  labs(
    title = "Basidiospore Length vs Width (with Q = 1 Reference Line)",
    x = "Width (µm)",
    y = "Length (µm)",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right"
  )


```


Building a graph that visualizes spore measurements per specimen as point and cross hairs
  - First need to oragnize basidiospore data and assign each specimen their species and section names
  - Organize each specimen dataframe into a list that is more manageable
  - Build crosshair plot color coded for section
  
Below code just helps to organize basidiospores_filled.csv
```{r}
#Assigning species names to taxa
setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore")

library(readr)
library(dplyr)
library(stringr)

# ---- Load specimen → species lookup ----
specimen_lookup <- read_csv("specimen_lookup.csv", show_col_types = FALSE)

# Ensure types are correct
specimen_lookup$specimen_id <- as.character(specimen_lookup$specimen_id)

# ---- Directory with spore CSVs ----
data_dir <- "basidiospore_by_specimen"

csv_files <- list.files(
  path = data_dir,
  pattern = "_basidiospore\\.csv$",
  full.names = TRUE
)

# ---- Container for data frames ----
spore_dfs <- list()

# ---- Explicit for-loop (robust and clear) ----
for (f in csv_files) {

  # Extract specimen ID from filename
  specimen_id <- str_extract(basename(f), "\\d+")

  # Read CSV
  df <- read_csv(f, show_col_types = FALSE)

  # Look up species
  species_name <- specimen_lookup$species[
    specimen_lookup$specimen_id == specimen_id
  ]

  # Defensive check
  if (length(species_name) == 0) {
    warning(paste("No species found for specimen", specimen_id))
    species_name <- NA_character_
  }

  # Add metadata
  df <- df %>%
    mutate(
      specimen_id = specimen_id,
      species = species_name
    )

  # Store in list, named by specimen_id
  spore_dfs[[specimen_id]] <- df
}

spore_all <- bind_rows(spore_dfs)

# We actually have to summarize the data for each of these specimens anyways which is going to be easier to do by using basidiospore_filled.csv which has min-10p-90p-max values but I realize they need to have averages as well. So we are going to compute average spore length and width for each specimen and add it to basidiospore_filled.csv


avg_spores <- spore_all %>%
  group_by(specimen_id) %>%
  summarise(
    mean_length = mean(Length, na.rm = TRUE),
    mean_width  = mean(Width,  na.rm = TRUE),
    .groups = "drop"
  )


```
  
Data set 'basidiospores_filled.csv' is complete and can now be used to generate 2 figures:
  1. Crosshairs plot visualizing each specimens spore shape measurement space
  2. Qvalue variation for specimens grouped by species and section
  
```{r}
setwd("~/Desktop/wiscam_microscopy/data/rt_basidiospore")
library(ggplot2)
library(readr)
library(dplyr)
library(stringr)

#Load .csv
spore_range <- read_csv("basidiospore_filled.csv", show_col_types = FALSE)


```

